---
title: "Introduction to multiple regression"
subtitle: "Notes and in-class exercises"
format: 
  html:
    embed-resources: true
    toc: true
---

```{r warning=FALSE, message=FALSE}
# Load packages
library(readr)
library(ggplot2)
library(dplyr)

# Load data
library(palmerpenguins)
data(penguins)
penguins <- penguins %>% 
  filter(species != "Adelie", bill_length_mm < 57)

# Check it out
head(penguins)
```

## Exercise 1: Visualizing the relationship

a. no wrong answer

b. There are multiple options!

```{r}
penguins %>% 
  ggplot(aes(y = flipper_length_mm, x = bill_length_mm, color = species)) + 
  geom_point()

penguins %>% 
  ggplot(aes(y = flipper_length_mm, x = bill_length_mm, shape = species)) + 
  geom_point()
```










## Exercise 2: Visualizing the model

a. no wrong answer

b. 

```{r}
penguins %>% 
  ggplot(aes(y = flipper_length_mm, x = bill_length_mm, color = species)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
```




## Exercise 3: Intuition

a. Gentoo tend to have longer flippers.
b. Flipper length is positively associated with bill length.
c. No. the lines are parallel / have the same slopes.









## Exercise 4: Model formula


a.  `bill_length_mm + species`
b. E[flipper_length_mm | bill_length_mm, speciesGentoo] = 127.75 + 1.40 * bill_length_mm + 22.85 * speciesGentoo

```{r}
# Build the model
penguin_mod <- lm(flipper_length_mm ~ bill_length_mm + species, data = penguins)

# Summarize the model
coef(summary(penguin_mod))
```





## Exercise 5: Sub-model formulas

Chinstrap: flipper_length_mm = 127.75 + 1.40 bill_length_mm

Gentoo:    flipper_length_mm = (127.75 + 22.85) + 1.40 bill_length_mm = 150.6 + 1.40 bill_length_mm





## Exercise 6: coefficients -- physical interpretation

a. The intercept coefficient, 127.75, is the intercept of the line for **Chinstrap** penguins.

b. The `bill_length_mm` coefficient, 1.40, is the **slope** of both lines.

c. The `speciesGentoo` coefficient, 22.85, indicates that the **intercept** of the line for Gentoo is 22.85mm higher than the **intercept** of the line for Chinstrap. Similarly, since the lines are parallel, the line for Gentoo is 22.85mm higher than the line for Chinstrap at any `bill_length_mm`.
    
    



## Exercise 7: coefficients -- contextual interpretation

a. For Chinstrap penguins with 0mm bills (silly), we expect a flipper length of 127.75mm.

b. For both Chinstrap and Gentoo penguins, flipper lengths increase by 1.40mm on average for every additional mm in bill length.

c. At any `bill_length_mm`, we expect the a Gentoo penguin to have 22.85mm longer flippers than a Chinstrap, on average.





## Exercise 8: Prediction

```{r}
# a
127.75 + 1.40*50 + 22.85*0

# b
127.75 + 1.40*50 + 22.85*1

# c
predict(penguin_mod,
        newdata = data.frame(bill_length_mm = 50, 
                             species = "Chinstrap"))
predict(penguin_mod,
        newdata = data.frame(bill_length_mm = 50, 
                             species = "Gentoo"))
```






## Exercise 9: R-squared

a. species
b. no wrong answer
c. It's higher than the R-squared when we use either predictor alone!

