---
title: "Solutions for: Multiple regression principles"
subtitle: "Notes and in-class exercises"
format: 
  html:
    embed-resources: true
    toc: true
---

<!-- NOTES to instructor: -->
<!-- - I recommend doing the first 3 exercises as a class. They're a review of the previous class. -->


# Notes

## Learning goals

Working with multiple predictors in our plots and models can get complicated!
There are no *recipes* for this process.
BUT there are some guiding *principles* that assist in long-term retention, deeper understanding, and the ability to generalize our tools in new settings.
By the end of this lesson, you should be familiar with some general *principles* for... 

- incorporating additional quantitative or categorical predictors in a visualization
- how additional quantitative or categorical predictors impact the physical representation of a model
- interpreting quantitative or categorical coefficients in a multiple regression model





## Readings and videos

Please watch the following video **before** class.

- [Interpreting multivariate models](https://www.youtube.com/watch?v=DnXwu1OWgMM) ([slides](https://drive.google.com/file/d/1fiJL1IbReg6RJmjxRideYzYTyQtWR3Or/view?usp=sharing))





# Exercises

Let's revisit the bikeshare data:

```{r warning = FALSE, message = FALSE}
# Load packages & import data
library(readr)
library(ggplot2)
library(dplyr)

bikes <- read_csv("https://mac-stat.github.io/data/bikeshare.csv") %>% 
  rename(rides = riders_registered)
```



## Exercise 1: Review visualization

```{r}
bikes %>% 
  ggplot(aes(y = rides, x = windspeed, color = weekend)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
```


## Exercise 2: Review model

weekdays: rides = 4738.38 - 63.97 windspeed    

weekends: rides = 4738.38 - 63.97 windspeed - 925.16 = 3813.22 - 63.97 windspeed


## Exercise 3: Review coefficient interpretation

a. We expect an average of 4738 riders on **weekdays with 0 windspeed**.

b. On both weekends and weekdays, we expect an average of 64 fewer riders for every 1mph increase in windspeed.

c. At any windspeed, we expect an average of 925 *fewer* riders on weekend days than on weekdays.




## Exercise 4: 2 categorical predictors – visualization


```{r}
bikes %>% 
  ggplot(aes(y = rides, x = season, fill = weekend)) + 
  geom_boxplot()
```




## Exercise 5: follow-up

a. In every season, ridership tends to be lower on weekends. Across weekend status, ridership tends to be highest in summer and lowest in winter.

b. 8 expected outcomes



## Exercise 6: 2 categorical predictors – build the model

a.

```{r}
#fall weekday:    
4260.45 - 912.33*0 - 116.38*0 + 438.44*0 - 1719.06*0

#winter weekday:
4260.45 - 912.33*0 - 116.38*0 + 438.44*0 - 1719.06*1

#fall weekend:    
4260.45 - 912.33*1 - 116.38*0 + 438.44*0 - 1719.06*0

#winter weekend:
4260.45 - 912.33*1 - 116.38*0 + 438.44*0 - 1719.06*1
```

b. 8: 2 weekend categories * 4 season categories    


## Exercise 7: 2 categorical predictors – interpret the model


a. We expect there to be an average of 4260 riders on **weekdays** during the **fall**.
b. *In any season*, we expect there to be, on average, 912 **fewer** riders on weekends than on **weekdays**.
c. *On both weekdays and weekends*, we expect there to be 1719 **fewer** riders in winter than in **fall**, on average.



## Exercise 8: 2 quantitative predictors – visualization


```{r}
bikes %>% 
  ggplot(aes(y = rides, x = windspeed, size = temp_feel)) + 
  geom_point(alpha=0.5) 

bikes %>% 
  ggplot(aes(y = rides, x = windspeed, color = temp_feel)) + 
  geom_point() 
```




## Exercise 9: follow-up

Ridership tends to increase with temperature (no matter the windspeed) and decrease with windspeed (no matter the temperature).



## Exercise 10: 2 quantitative predictors – modeling

a. -24.06 = average expected ridership on days with 0 windspeed and a 0 degree temperature. (This obviously doesn't make contextual sense--how could we make it more interpretable?)

b. No matter the temperature, we expect 37 fewer riders for every 1mph increase in windspeed.

c. No matter the windspeed, we expect 56 more riders for every 1 degree increase in temperature.




## Exercise 11: Which is best?

a. model 2
b. temperature
c. R-squared increases (our model is stronger when we include another predictor)
d. nope. model 1 (with windspeed and weekend) has a lower R-squared than model 5 (with only temperature)
    

\
\


## Exercise 13: Practice 1    

```{r}
bikes %>% 
  ggplot(aes(y = rides, x = weekend, fill = weather_cat)) + 
  geom_boxplot()

new_model_1 <- lm(rides ~ weekend + weather_cat, bikes)
coef(summary(new_model_1))
```



- The average expected ridership on a *weekday* with *nice* weather (categ1) is 4212 rides.    

- On days with similar weather, the average expected ridership on a *weekend* is 982 rides less than on weekdays.    

- On similar days of the week, the average expected ridership when the weather is "dreary" (categ2) is 609 rides less than when the weather is nice.    



\
\




## Exercise 14: Practice 2    
    
```{r}
bikes %>% 
  ggplot(aes(y = rides, x = temp_feel, color = humidity)) + 
    geom_point()

new_model_2 <- lm(rides ~ temp_feel + humidity, bikes)
coef(summary(new_model_2))
```

- It doesn't really make sense to interpret the intercept -- we didn't see any days that were 0 degrees with 0 humidity.    

- On days with the same humidity level (or "holding humidity constant"), ridership is expected to increase by 60 rides for each 1 degree increase in temperature.    

- On days with the same temperature (or "holding temperature constant"), ridership is expected to decreases by 1867 * 0.1 = 187 rides for each 0.1 increase in humidity levels.    



\
\




## Exercise 15: Practice 3

```{r}
new_model_3 <- lm(rides ~ temp_feel + weather_cat, bikes)
coef(summary(new_model_3))

bikes %>% 
  ggplot(aes(y = rides, x = temp_feel, color = weather_cat)) + 
  geom_point() + 
  geom_line(aes(y = new_model_3$fitted.values), size = 1.5)
```

- It doesn't really make sense to interpret the intercept -- we didn't see any days that were 0 degrees.    

- On average, assuming the weather is the same, ridership is expected to increases by 55 rides for each 1 degree increase in temperature.    

- On average, ridership is expected to decrease by 386 rides on a dreary weather day (categ2) than a nice weather day (categ1), assuming the temperature is the same.


## Exercise 16: CHALLENGE

   
```{r}
bikes %>% 
  ggplot(aes(y = rides, x = temp_feel, color = weekend, size = humidity)) + 
  geom_point()

new_model_4 <- lm(rides ~ temp_feel + humidity + weekend, bikes)
coef(summary(new_model_4))
```

- It doesn't really make sense to interpret the intercept -- we didn't see any days that were 0 degrees.    
- Assuming humidity and time of week are the same, ridership is expected to increase by 59 rides for each 1 degree increase in temperature.    
- Assuming humidity and time of week are the same, ridership is expected to decrease by 1906*0.1 = 190.6 rides for each 0.1 point increase in humidity levels.       
- Holding humidity and temperature constant, ridership is expected to be 869 rides lower on weekends vs weekdays. 


This model would look like 2 parallel planes, one for weekends and one for weekdays! Why?  `rides` vs `temp_feel` (quant) would be a line. Adding in `humidity` (quant) would turn the model into a plane. Adding in `weekend` (cat) would split this one plane into two unique planes, one for each weekend category!



