---
title: "Solutions for Multiple linear regression: exploring interaction"
subtitle: "Notes and in-class exercises"
format: 
  html:
    embed-resources: true
    toc: true
---

# Notes

## Learning goals

By the end of this lesson, you should be able to:

- Describe when it would be useful to include an interaction term to a model
- Write a model formula for an interaction model
- Interpret the coefficients in an interaction model in the data context

## Readings and videos

Today is a day to discover ideas, so no readings or videos to go through before class.

<!-- This section is for any mini-lecture, review material, material to motivate today's lesson, more guided / structured examples than the exercises section below. -->



**File organization:** Save this file in the "Activities" subfolder of your "STAT155" folder.

# Class exploration

> **Guiding question:** What job sectors have the highest return on education?

We'll use data from the 2018 Current Population Survey to explore. The codebook for this data is available [here](https://mac-stat.github.io/data/cps_2018_codebook.html). For now we'll focus on individuals who have jobs in the management or transportation sectors to simplify our explorations.

```{r}
# Load packages and data
library(readr)
library(ggplot2)
library(dplyr)
cps <- read_csv("https://mac-stat.github.io/data/cps_2018.csv")

# Get data on just the management and transportation sectors
cps_sub <- cps %>% 
    filter(industry %in% c("management", "transportation"))
```


## Exercise 1: Wages across all industries

The plot below illustrates the relationship between wage and education for all of the industries in our `cps` dataset.

```{r}
# Plot
ggplot(cps, aes(y = wage, x = education, color = industry)) + 
    geom_smooth(method = "lm", se = FALSE)
```

a. The industry-specific lines all have different slopes.

b. ag (first in alphabetical order)

c. This is a challenge! Compare your prediction to what you see when fitting the model in part d.

d. 

```{r}
# Fit an interaction model called wage_model
wage_model <- lm(wage ~ education*industry, data = cps)

# Display summarized model output
coef(summary(wage_model))
```

e. In the **management** industry, wages increase the *most* per year of education. The increase is 61.95396 + 8616.31984 = \$8678.274 per year. That is, every additional year of education is associated with an average increase of \$8678.27 in yearly wages in the management industry.

f. In the **ag**riculture industry, wages increase the *least* per year of education. The increase is \$61.95 per year.  That is, every additional year of education is associated with an average increase of \$61.95 in yearly wages in the ag industry.


## Exercise 2: Thinking beyond

If a variable `x` has an interaction with the `industry` variable in affecting wages, then the relationship between `x` and wages must be different by industry. We might suspect that this could be the case for `hours` worked per week. We can make a plot to verify that this is actually the case:

```{r}
ggplot(cps, aes(y = wage, x = hours, color = industry)) + 
    geom_smooth(method = "lm", se = FALSE)
```


