---
title: "Solutions for 20. Bootstrapping"
subtitle: "Notes and in-class exercises"
format: 
  html:
    embed-resources: true
    toc: true
---


# Notes

You will not be able to render this document until you've filled in the code!



## Learning goals

Let $\beta$ be some *population parameter* and $\hat{\beta}$ be a *sample estimate* of $\beta$.
In order to study the potential error in $\hat{\beta}$, you will...

- explore two approaches to approximating the sampling distribution of $\hat{\beta}$:       
    - Central Limit Theorem (CLT)
    - bootstrapping

- identify the difference between *sampling* and *resampling*

- intuit how bootstrapping results can be used to make inferences about $\beta$








## Readings and videos

Please watch the following video **after** class:

- [bootstrapping](https://voicethread.com/myvoice/thread/15611638/97563140)














```{r}
# Load the data & packages
library(tidyverse)
library(readr)
fish <- read_csv("https://Mac-STAT.github.io/data/Mercury.csv")
```

Plot and model the relationship of mercury concentration with length:

```{r}
fish %>% 
  ggplot(aes(y = Concen, x = Length)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)

fish_model <- lm(Concen ~ Length, data = fish)
coef(summary(fish_model))
```


## Exercise 1: sample vs population

a. sample estimate

b. no idea (at this point at least)! we don't know the actual population slope





## Exercise 2: The rub

We only have access to a *sample* of fish, not the entire population.








## Exercise 3: CLT

0.005






## Exercise 4: Challenge


The results are all the same.
Since we sampled without replacement, we got the same sample hence the same sample estimates every time.





## Exercise 5: Resampling

a. We always get the same sample (the original 5 fish), just in a different order. Thus the mean is always the sample.

b. They differ.

c. Sampling the same fish more than once is like observing different fish with similar characteristics.










## Exercise 6: Bootstrapping


```{r}
# Set the seed so we get the same results
set.seed(155)

# Build 500 bootstrap models using REsamples of size 171
fish_models_bootstrap <- mosaic::do(500)*(
  fish %>% 
    sample_n(size = 171, replace = TRUE) %>% 
    with(lm(Concen ~ Length))
)

head(fish_models_bootstrap)
```






## Exercise 7: Bootstrap results


a. `fish_model`

b. intuition

c. They are normally distributed around our original sample estimate.








## Exercise 8: Bootstrap standard errors


a. .

```{r}
fish_models_bootstrap %>%
  summarize(sd(Length))
```


b. very close!


  
    
    
## Exercise 9: Looking ahead at intervals


a. Will vary. Maybe 0.0375ish to 0.07ish.



b. Will vary. For example, if we lop off the extreme 2.5% (the smallest and biggest estimates) and just keep the middle 95%:

```{r}
fish_models_bootstrap %>%
  summarize(quantile(Length, 0.025), quantile(Length, 0.975))
```







## Exercise 10: Looking ahead at hypothesis testing


a. Yes. All of the lines reflect a positive association.


b. Yes. It falls entirely above 0 (a slope of 0 indicating no association).



